\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-23 chapters/bci.tex
started wcble introduction

The major drawback of \ac{eeg} is its poor spatial resolution.
The electrodes are large and sit on the outside of the scalp, relatively far
away from the cortex.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
The noise problem mentioned earlier is also prominently present in \ac{eeg}
recordings.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the environment, the power
supply, nearby electrical equipment, or the user's muscle movements.
Combined, this means \ac{eeg} is inherently noisy.

@@

The \ac{bci} paradigm~\cite{Xu2021} is the key to translating brain activity into useful
output commands, since it defines which neural responses will be elicited and
captured.
A paradigm can be loosely seen as \emph{a way of interacting with the
\ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on
external stimulation and the engagement of the user as shown in
Figure~\ref{fig:bci-paradigms}.

\fullpagefig{figures/bci/paradigms.eps}{
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in \emph{active}, \emph{reactive} and \emph{passive}
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{fig:bci-paradigms}

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as passive
\ac{bci}.
From the user's point of view, this concept is probably the most attractive,
since their intentions or cognitive state state would be inferred from just
their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.

Furthermore, active paradigms make it easier to quickly collect training data for
supervized machine learning since the conditions can be more easily controlled
by the user or the stimulation.

active

reactive
works for most people
high effort, unintuitive and low expressive power compared to movement and
speech~\cite{Zander2011}, which are more natural and can be emulated with active bci

\todo{check permissions}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback
error potentials
cannot easily be used for fast communication
synchronous <-> asynchronous
Again, asynchronous is preferred by users because it is more intuitive and
comfortable to use, but it is far easier to decode synchronous responses since
it is known when to expect a response and when a decision should be made.

active:
motor imagery
imagined speech

reactive:

somatosensory evoked potentials~\cite{Petit2021} (cite jimmy)

auditory:
ERP/oddball, steady state

visual paradigms:
more performant since it allows us to use our visual system, the most adapted
sensory system in humans
SSVEP, cVEP, mVEP, VEP/P300/Oddball
high ssvep itr

hybrid

@@

\section{A case study: the visual oddball ERP paradigm}
good choice because it scores high on the parameters above
\subsection{Stimulation}
comfortable, preferred over SSVEP, similar ITR is this true?
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}
\subsection{The event-related potential}
P300 and other components
does not only rely on P300, but also heavily dependent on visual early
components, hence we'll refer to it under the more generall oddball ERP
paradigm
\todo{figure of ERP components}
\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques

>>>>

The major drawback of \ac{eeg} is its poor spatial resolution.
The electrodes are large and sit on the outside of the scalp, relatively far
away from the cortex.
Furthermore, the layers of skull and cerebrospinal fluid in between can
attenuate the signal and cause volume conduction, which results in electrodes
measuring a mixture of activity from sources elsewhere in the brain.
This also has a negative effect on spectral resolution.
The noise problem mentioned earlier is also prominently present in \ac{eeg}
recordings.
In addition to picking up brain activity other than that of the region of
interest, \ac{eeg} also records other artifacts, from the environment, the power
supply, nearby electrical equipment, or the user's muscle movements.
Combined, this means \ac{eeg} is inherently noisy.
\todo{citations from 'Decoding covert speech from eeg-a cromprehensive review
EEG has lower signal-to-noise ratio (SNR) than the other
modalities. It is almost always corrupted by artifacts such as
muscular artifacts (Eberle et al., 2012; Liu, 2019).
• EEG has limited spectral and spatial resolution (Peled et al.,
2001; Lakshmi et al., 2014).
• Recording EEG for longer duration is challenging since the
conductive gel or the saline solution applied for reducing the
electrode impedance dries up over time, thus increasing the
electrode impedance (Guger et al., 2012; Xu et al., 2018a).
• A trained personnel is required for placing the EEG
electrode cap}

@@

The \ac{bci} paradigm~\cite{Xu2021,Neeling2019} is the key to translating brain activity into useful
output commands, since it defines which neural responses will be elicited and
captured as features.
A paradigm can be loosely seen as \emph{a way of interacting with the
\ac{bci}}.
According to \textcite{Zander2011}, they can be arranged by their reliance on
external stimulation and the engagement of the user as shown in
Figure~\ref{fig:bci-paradigms}.

\fullpagefig{figures/bci/paradigms.eps}{
  A \ac{bci} paradigm defines which brain activity will be used for control.
  They can be categorized in active, reactive and passive
  paradigms, depending on their reliance on external stimulation and active
  participation required from the user.
  Adapted from~\textcite{Muehl2014}.
}{fig:bci-paradigms}
\todo{check permissions}

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as \emph{passive
\ac{bci}}.
From the user's point of view, this concept is probably the most attractive,
since their intentions or cognitive state state would be inferred from just
their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
Furthermore, active paradigms make it easier to quickly collect training data for
supervized machine learning since the conditions can be more easily controlled
by the user or the stimulation.

Paradigms with high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{
  \cite{Becker2022,Thompson2019}
}.
\todo{footnote on how this just means it needs to be more adapted,
inclusivity, harmful to progress, paraphrase abstracts of articles}

Reactive \ac{bci} take another approach, by giving the user a discrete set of
sensory stimulations and decoding towards which stimulus the user's attention
was directed.
Compared to active \ac{bci}, reactive paradigms usually require higher effort
and can induce fatigue due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g. speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

Reactive \ac{bci} can be realized in multiple ways, depending on the
stimulation used.
Some examples include the following:
In the steady-state somatosensory evoked potentials paradigm~\cite{Petit2021},
the user can attend to one of multiple vibrotactile stimulations in different
limbs, which encodes the information of the attended limb in the brain signal.
In auditory paradigms, information can be modulated by the volume, tone, pitch,
spatial origin of presented sounds, to which the user can
attend~\cite{Kaongoen2017}.
Most common are the visual paradigms.
These can be more performant since they allows the user to use their visual
system, one of the most adapted sensory system in humans.

The major visual paradigms comprise
\ac{ssvep}, \ac{cvep}, oddball and \ac{mvep} type interfaces.
In \ac{ssvep}, information is modulated by the frequency of a stimulus attended
among many that each flicker continuously at different
frequencies~\cite{Chen2021}.
In \ac{cvep}, the stimuli flicker instead with distinct on-of patterns, which can
be correlated with the brain activity to retrieve the attended stimulus.
Instead of stimulating all possible selections at once, the oddball paradigm
stimulates them one by one with single flashing.\todo{reference}
Since the times of stimulation of each target is known by the interface,
information is modulated by time of stimulation, and times where an attention
response can be found correspond to selected targets.
\Ac{mvep} is usually similarly time modulated, but stimuli make sudden
movements in specific directions instead of flashing.
Information is then also carried by movement
direction~\cite{Libert2021a,Libert2022}.

The \ac{bci} paradigms mentioned above can also be combined to gain performance.
Straightforward examples are activating or deactivating \ac{ssvep} stimulation
with a motor command~\cite{Neeling2019}, or combining multiple visual paradigms by stimulating
along multiple dimensions at once.
\textcite{Han2023} recently used this strategy, combining frequancy and phase
coding in \ac{ssvep} with the \ac{mvep} and oddball paradigms to develop an
efficient \ac{bci} where one of 200 targets can be accurately selected with
only 800ms of stimulation.
Finally, paradigms can be implemented in a \emph{synchronous} or
\emph{asynchronous} manner.
Asynchronous \ac{bci} constantly decodes the signal looking for the response of
interest, while in synchronous \ac{bci} decisions are made (and eventual
stimulation takes place) at predetermined time points.
Again, asynchronous is preferred by users because it is more intuitive and
comfortable to use, but it is far easier to decode synchronous responses since
it is known when to expect a response and if a decision should be made.

@@

\section{A case study: the visual oddball paradigm}

The visual oddball paradigm is a reactive, stimulus-dependent, synchronous
\ac{bci} paradigm.
good choice because it scores high on the parameters above
well established, good research, first application cite
Flashing in various configurations, such as row-colum or in groups,

\subsection{Stimulation}
comfortable, preferred over SSVEP, similar ITR is this true?
\todo{figure of P3 matrix}
\todo{cite P300 literature, first example and current sota}

\subsection{The event-related potential}

Because of this, the oddball paradigm is often referred to as the P300
paradigm.
However, due to equal importance of other \ac{erp} components than the P3 in
decoding as we will show later, this could be considered a bit of a misnomer
and we will stick to the oddball paradigm naming.

P300 and other components, explain function (Luck)

\todo{figure of ERP components}

\subsubsection{P1}
\subsubsection{N1}
\subsubsection{P2}
\subsubsection{N2}
\subsubsection{P3}
\subsubsection{Late negativity}

\subsection{Preprocessing and feature extraction}
referencing, filtering, epoching, baselining
\subsection{Decoding ERPs}
tLDA, beamforming
Linear decoding techniques
